---
title: "Problem-Set-1 for POLI 271"
format: pdf
editor: visual
---

# 1. Univariate displays & sampling distributions

## *Problem a*
```{r, fig.width=7, fig.height=5}
s_values <- c(10, 100, 1000)
n_values <- c(10, 100, 1000)

par(mfrow = c(3, 3))
for (s in s_values) {
  for (n in n_values) {
    sample_means <- replicate(s, mean(sample(1:20, n, replace = TRUE)))

    hist(sample_means,
         breaks = 20,
         col = "grey",
         border = "black",
         main = paste("s =", s, ", n =", n),
         xlab = "",
         xlim = c(1, 20))
    
    abline(v = mean(sample_means), col = "red", lwd = 0.5)
  }
}


```
## *Problem b*
The histograms shows that when increasing the sample size reduces variability in sample means. I set it to 1:20 so it's getting closer to 10. And increasing the number of samples makes it look like normal distribution. The key assumption here is what CLT describes: the distribution of a normalized version of the sample mean converges to a standard normal distribution. 

# 2. Monte Carlo integration

```{r}
func = function(x){exp(-x)*sin(x)}
result = integrate(func, lower = 2, upper = 5)
print(result)
```
# 3. Systematic and stochastic components

## *Problem a*

Systematic Component: $y_i=1+0.5x_{i1}-2.2x_{i2}+x_{i3}$
Stochastic Component: ${\epsilon_i}\sim{N(\mu=0,\sigma^2=1.5)}$

## *Problem b*
**Part I.** The dimensions of $\mathbf{X}$ is denoted as $n$ is $2$. 
```{r}
data = read.csv("xmat.csv")
print(dim(data))
head(data, 10)
```
**Part II.**

```{r}
set.seed(10825) 
# Why not 42 and 3407
coefficient_0 = 1
coefficient_1 = 0.5
coefficient_2 = -2.2
coefficient_3 = 1
x_1 = data$X1
x_2 = data$X2
x_3 = data$X3
e = rnorm(n = nrow(data), mean = 0, sd = sqrt(1.5))
y = coefficient_0 + 
  coefficient_1*x_1 + coefficient_2*x_2 + coefficient_3*x_3 + e

linear = lm(y ~ x_1 + x_2 + x_3)
summary((linear))
```
# 4. OLS in matrix form
