---
title: "POLI 271 Problem set 4"
format: pdf
editor: visual
---

# POLI 271 Problem set 4

## Problem 1

### a

```{r}
library(tidyverse)
library(gt)
library(modelsummary)
library(stargazer)
library(ROCR)
library(caret)
library(cvTools)
library(MASS)
options(warn = -1)
options(message = -1)
```


```{r}
flmdw <- read.csv("flmdw-1.csv")
head(flmdw, 20)
```
```{r}
library(ggplot2)
ggplot(flmdw, aes(x = factor(onset))) +
  geom_bar(fill = c("violet", "purple")) +
  labs(title = "Distribution of Civil War Onset",
       x = "Civil War Onset",
       y = "Count") +
  scale_x_discrete(labels = c("No Onset", "Onset")) +
  theme_minimal()

```
This is a rare event. Bayesian Logistic Regression; LASSO

## b

```{r}
library(dplyr)
colnames(flmdw)
flmdw_complete <- flmdw %>%
  dplyr::select(onset, gdpenl, lpopl1, lmtnest, Oil, polity2l, relfrac) %>%
  na.omit()

model1 <- glm(onset ~ gdpenl + lpopl1 + lmtnest, 
              data = flmdw_complete, family = binomial)
model2 <- glm(onset ~ gdpenl + lpopl1 + lmtnest, 
              data = flmdw_complete, family = binomial(link = "probit"))
model3 <- glm(onset ~ gdpenl + lpopl1 + lmtnest + Oil + polity2l + relfrac, 
              data = flmdw_complete, family = binomial(link = "probit"))
model4 <- glm(onset ~ gdpenl + lpopl1 + lmtnest + Oil + polity2l * relfrac, 
              data = flmdw_complete, family = binomial(link = "probit"))

summary(model1)
summary(model2)
summary(model3)
summary(model4)

library(broom)
glance(model1)
glance(model2)
glance(model3)
glance(model4)
```
**Based on the results and AIC values, Model 4 is the most promising due to its lower AIC and inclusion of interaction effects, which likely capture more nuanced relationships between variables.**

```{r}
library(broom)
library(gt)

model_summaries <- bind_rows(
  tidy(model1, conf.int = TRUE) %>% mutate(model = "Model 1"),
  tidy(model2, conf.int = TRUE) %>% mutate(model = "Model 2"),
  tidy(model3, conf.int = TRUE) %>% mutate(model = "Model 3"),
  tidy(model4, conf.int = TRUE) %>% mutate(model = "Model 4")
)

gt_table <- gt(model_summaries) %>%
  tab_header(
    title = "Results of Logistic and Probit Regression Models"
  ) %>%
  fmt_number(
    columns = vars(estimate, std.error, statistic, p.value, conf.low, conf.high),
    decimals = 3
  ) %>%
  cols_label(
    estimate = "Estimate",
    std.error = "Std. Error",
    statistic = "z value",
    p.value = "P Value",
    conf.low = "CI Low",
    conf.high = "CI High",
    term = "Term"
  ) %>%
  tab_spanner(
    label = "Confidence Interval",
    columns = vars(conf.low, conf.high)
  )

gtsave(gt_table, filename = "gt_table.pdf")
```
## c

```{r}
pred1 <- prediction(predict(model1, type = "response"), flmdw_complete$onset)
roc1 <- performance(pred1, "tpr", "fpr")

pred2 <- prediction(predict(model2, type = "response"), flmdw_complete$onset)
roc2 <- performance(pred2, "tpr", "fpr")

pred3 <- prediction(predict(model3, type = "response"), flmdw_complete$onset)
roc3 <- performance(pred3, "tpr", "fpr")

pred4 <- prediction(predict(model4, type = "response"), flmdw_complete$onset)
roc4 <- performance(pred4, "tpr", "fpr")

plot(roc1, col = "blue", main = "ROC Curves Comparison", xlim = c(0, 1), ylim = c(0, 1))
plot(roc2, col = "red", add = TRUE)
plot(roc3, col = "green", add = TRUE)
plot(roc4, col = "purple", add = TRUE)
legend("bottomright", legend = c("Model 1", "Model 2", "Model 3", "Model 4"),
       col = c("blue", "red", "green", "purple"), lty = 1)
```
## d

```{r}
model_restricted <- glm(onset ~ gdpenl + lpopl1 + lmtnest + Oil + relfrac, 
                        data = flmdw_complete, family = binomial(link = "probit"))

lr_test <- anova(model_restricted, model4, test = "Chisq")

print(lr_test)

```
The p-value (0.02528) is less than 0.05, meaning we reject the null hypothesis that $β_{dem} = β_{demfrac}$ = 0

## e

```{r}
library(caret)
library(pROC)

set.seed(3407) # 3407 is all you need

flmdw_complete$onset <- factor(flmdw_complete$onset, levels = c(0,1), labels = c("No", "Yes"))

cv_control <- trainControl(method = "cv", number = 10, classProbs = TRUE, summaryFunction = twoClassSummary)

cv_model1 <- train(onset ~ gdpenl + lpopl1 + lmtnest, 
                   data = flmdw_complete, method = "glm", family = binomial(), 
                   trControl = cv_control, metric = "ROC")

cv_model2 <- train(onset ~ gdpenl + lpopl1 + lmtnest + Oil, 
                   data = flmdw_complete, method = "glm", family = binomial(), 
                   trControl = cv_control, metric = "ROC")

cv_model3 <- train(onset ~ gdpenl + lpopl1 + lmtnest + Oil + polity2l + relfrac, 
                   data = flmdw_complete, method = "glm", family = binomial(link = "probit"), 
                   trControl = cv_control, metric = "ROC")

cv_model4 <- train(onset ~ gdpenl + lpopl1 + lmtnest + Oil + polity2l * relfrac, 
                   data = flmdw_complete, method = "glm", family = binomial(link = "probit"), 
                   trControl = cv_control, metric = "ROC")

prob1 <- predict(cv_model1, flmdw_complete, type = "prob")[,"Yes"]
prob2 <- predict(cv_model2, flmdw_complete, type = "prob")[,"Yes"]
prob3 <- predict(cv_model3, flmdw_complete, type = "prob")[,"Yes"]
prob4 <- predict(cv_model4, flmdw_complete, type = "prob")[,"Yes"]

roc1 <- roc(flmdw_complete$onset, prob1)
roc2 <- roc(flmdw_complete$onset, prob2)
roc3 <- roc(flmdw_complete$onset, prob3)
roc4 <- roc(flmdw_complete$onset, prob4)

plot(roc1, col = "blue", main = "10-Fold Cross-Validation ROC")
lines(roc2, col = "red")
lines(roc3, col = "green")
lines(roc4, col = "purple")
legend("bottomright", legend = c("Model 1", "Model 2", "Model 3", "Model 4"),
       col = c("blue", "red", "green", "purple"), lty = 1)
auc(roc1)
auc(roc2)
auc(roc3)
auc(roc4)
```
**Clearly model4 has the highest AUC. Hard to reject this one. **

## f

```{r}
# Prepare data for predictions
prediction_data <- expand.grid(
  polity2l = seq(min(flmdw_complete$polity2l), max(flmdw_complete$polity2l), by = 1),
  relfrac = c(0.1, 0.5, 0.9),
  gdpenl = mean(flmdw_complete$gdpenl), 
  lpopl1 = mean(flmdw_complete$lpopl1),
  lmtnest = mean(flmdw_complete$lmtnest),
  Oil = mean(flmdw_complete$Oil)
)

# Predict probabilities
prediction_data$predicted_prob <- predict(cv_model4, newdata = prediction_data, type = "prob")[, "Yes"]

# Plotting
ggplot(prediction_data, aes(x = polity2l, y = predicted_prob, color = factor(relfrac))) +
  geom_line() +
  labs(x = "Democracy Index", y = "Probability of Civil War Onset") +
  scale_color_brewer(palette = "Set1") +
  theme_minimal()
```
The data suggest that while democracy is generally promoted as a means to prevent conflict, its effectiveness can vary significantly depending on the religious composition of a society. In highly diverse societies, the introduction of democracy should be handled with care, potentially supplemented by measures that promote intergroup dialogue and reconciliation to mitigate the risks of increased conflict.